---
title: "HW3"
author: "James Dalgleish"
date: "October 9, 2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
```
First, we'll begin by importing the dataset, cleaning the names to be more appropriate, and
focusing on the Overall Health topic (by filtering for this topic only in the BRFSS responses).
Responses are filtered to only include the 5 ordinal categories, Excellent, Very Good,
Good, Fair, Poor.
```{r data_import}
brfss <- p8105.datasets::brfss_smart2010 %>% #Pulls dataframe out of package.
  janitor::clean_names() %>%  #Converts to snake case.
  filter(topic == "Overall Health",
         response %in% c("Excellent","Very Good", "Good", "Fair", "Poor")) %>%     #Filters by overall health topic.
  mutate(response = factor(response,levels = c("Excellent","Very Good", "Good", "Fair", "Poor")))
  
  
```
Problem 2
```{r instacart_import}
instacart <- p8105.datasets::instacart %>% #Pulls dataframe out of package.
  janitor::clean_names() #cleans names
```
```{r distinct_aisles}
distinct_aisles <- instacart %>%
  group_by(aisle) %>% 
  summarize(count = n()) %>% 
  arrange(
    desc(count)
  )

```
There appear to be `r nrow(distinct_aisles)` aisles. The aisles with the most are displayed in the
table below:
```{r}
distinct_aisles %>% 
  head() %>% 
  kable()
```

The numbers of items in each aisle can be visualized with the following plot:

```{r}
items_by_aisle <- instacart %>%
  group_by(aisle) %>% 
count() %>% 
  arrange(
    desc(n)
    ) %>%
as.data.frame() %>% #makes the grouping variable modifyable 
  as.tibble() %>%  #factor reorder.
  mutate(aisle = forcats::fct_reorder(aisle, n))


items_by_aisle %>% 
  ggplot(aes(x = aisle,y = n,color = n)) + #establishes mappings.
  geom_col() + #like geom_bar, but takes a variable for height.
  scale_color_viridis_c() + #adds color
  scale_y_continuous(expand=c(0,0)) + #removes whitespace at bottom.
   theme_bw() +
  theme(axis.title.x = element_blank(), #removes panel grid at back.
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  labels(fill="Items in aisle") +
  ylab("Items in aisle") +
  xlab("aisles, sorted by number of items")
```

Now, we'll create a table with the most popular items in the aisles for
baking ingredients, dog food care, and packaged vegetables/fruits.
```{r}
instacart %>% 
  filter(aisle
         %in% c("baking ingredients","dog food care","packaged vegetables fruits")) %>% 
  group_by(aisle) %>% 
count(product_name) %>% 
  filter(n == max(n)) %>%
  arrange(
    desc(n)
  ) %>% 
  select(product_name,n,everything())

```

Instruction"Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table)."
Now, we'll show the mean hour of the day at which the items are ordered.
```{r}
pink_ice_tbl<-instacart %>% 
  filter(product_name %in% c("Pink Lady Apples","Coffee Ice Cream")) %>% 
  group_by(product_name,order_dow) %>% 
  summarize(mean_hour=mean(order_hour_of_day)) %>% 
  spread(key=product_name,value=,mean_hour) %>% 
mutate(order_dow = recode(.$order_dow,"0" = "Sunday",
                             "1" = "Monday",
                             "2" = "Tuesday",
                             "3" = "Wednesday",
                             "4" = "Thursday",
                             "5" = "Friday",
                             "6" = "Saturday")) %>% 
  janitor::clean_names() %>% 
  select(order_dow,
         coffee_ice_cream,
         pink_lady_apples)
pink_ice_tbl
```

